Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
Rough author contrib. feat imp limits

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are sparse, as these patients were likely on
antibiotics prior to CDI onset and may have begun antibiotic treatment prior to
the initial stool sample collection.

@@

<!--
# Acknowledgements

TODO funding & author contributions
-->

>>>>

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are sparse, as these patients were likely on
antibiotics prior to CDI onset and may have begun antibiotic treatment prior to
the initial stool sample collection.
A limitation of permutation importance is that the contribution of each feature
is considered in isolation, but members of microbial communities interact and
compete with each other, thus these complicated relationships are not well 
captured by permutation importance.

@@

# Acknowledgements

We thank the patients for donating stool samples and the research
team members that who collected, stored, and processed the samples.

<!--
## Author contributions

KLS performed data processing, trained machine learning models, wrote the
analysis code, created the figures and tables, and wrote the original draft of
the manuscript.
SET processed samples, performed initial analysis of IDSA severity scores, and
contributed analysis code.
MLC contributed analysis code and assisted in training machine learning models.
JW, VBY, and KR directed the analysis of clinical value.
PDS conceived of the study and supervised the project.
All authors reviewed and edited the manuscript.

## Funding

TODO

-->
